---
title: "Smoking and Blood Pressure"
author: "Matthew Poole Chicano"
date: "2022-07-27"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Prerequisites 

Loading tidyverse: 
```{r}
library(tidyverse)
library(DescTools)
```

Loading the dataset:
```{r}
heart <- read.csv("heart_cleaned.csv")
```


## Manipulating the Data

Viewing the structure:
```{r}
str(heart)
```

Removing unnecessary columns:
```{r}
heart <-
  select(
    heart,
    randid,
    sex,
    age,
    sysbp,
    diabp,
    cigpday,
    heartrte,
    prevhyp,
    hyperten,
    timehyp
  )

head(heart)
```


Filtering for smokers:
```{r}
smokers <-
  filter(
    heart,
    cigpday >= 1
  )

head(smokers)
```


Creating light, moderate, and heavy smokers group:
```{r}
smokers <-
  mutate(
    smokers,
    smkerst = 
      case_when(
        cigpday <= 5 ~ "light",
        cigpday >= 6 & cigpday <= 20 ~ "moderate",
        cigpday >= 21 ~ "heavy"
      )
  )

head(smokers, 10)
```


Converting necessary data into factors:
```{r}
smokers <- mutate(smokers, 
                  sex = factor(smokers$sex, labels = c("Male", "Female")),
                  prevhyp = factor(smokers$prevhyp, labels = c("free", "prev")),
                  hyperten = factor(smokers$hyperten, labels = c("free", "prev")),
                  smkerst = factor(smokers$smkerst, levels = c("light", "moderate", "heavy"),
                                                    labels = c("Light", "Moderate", "Heavy")))

str(smokers)
```

## Descriptive Stats

Comparing by smoker status:
```{r}
smokers_descrip <- summarise(group_by(smokers, smkerst), 
                             "N" = n(),
                             "Mean_SysBP" = mean(sysbp, na.rm = T),
                             "Mean_DiaBP" = mean(diabp, na.rm = T),
                             "Mean_Heartrate" = mean(heartrte, na.rm = T),
                             "SD_SysBP" = sd(sysbp, na.rm = T),
                             "SD_DiaBP" = sd(diabp, na.rm = T),
                             "SD_Heartrate" = sd(heartrte, na.rm = T),
                             "SE_SysBP" = sd(sysbp, na.rm = T)/sqrt(length(smkerst)),
                             "SE_DiaBP" = sd(diabp, na.rm = T)/sqrt(length(smkerst)),
                             "SE_Heartrate" = sd(heartrte, na.rm = T)/sqrt(length(smkerst)))

smokers_descrip
  
```


Comparing by smoker status x sex:
```{r}
smokers_descrip_sex <- summarise(group_by(smokers, smkerst, sex), 
                             "N" = n(),
                             "Mean_SysBP" = mean(sysbp, na.rm = T),
                             "Mean_DiaBP" = mean(diabp, na.rm = T),
                             "Mean_Heartrate" = mean(heartrte, na.rm = T),
                             "SD_SysBP" = sd(sysbp, na.rm = T),
                             "SD_DiaBP" = sd(diabp, na.rm = T),
                             "SD_Heartrate" = sd(heartrte, na.rm = T),
                             "SE_SysBP" = sd(sysbp, na.rm = T)/sqrt(length(smkerst)),
                             "SE_DiaBP" = sd(diabp, na.rm = T)/sqrt(length(smkerst)),
                             "SE_Heartrate" = sd(heartrte, na.rm = T)/sqrt(length(smkerst)))

smokers_descrip_sex
```


Summary of participants:
```{r}
summarise(group_by(smokers, sex), 
          "N" = n(),
          "Mean_Age" = mean(age),
          "SD" = sd(age),
          "Min" = min(age),
          "Max" = max(age))

```

Total:
```{r}
summarise(smokers,
          "N" = n(),
          "Mean_Age" = mean(age, na.rm = T),
          "SD" = sd(age, na.rm = T),
          "Min" = min(age, na.rm = T),
          "Max" = max(age, na.rm = T))
```

Group x sex x age:
```{r}
summarise(group_by(smokers, smkerst, sex),
          "N" = n(),
          "Mean_Age" = mean(age, na.rm = T),
          "SD" = sd(age, na.rm = T))
```


## Visualizing the Data

Bar graph of blood pressure:
```{r}
# creating new data frame to plot both SysBP and DiaBP into the same graph
smokers_tb <- tribble(~smkerst, ~BPtype, ~value, ~SE,
                      "Light", "SysBP", 135.0591, 2.5228948,
                      "Light", "DiaBP", 79.69355, 1.2241013,
                      "Moderate", "SysBP", 131.7338, 1.5415797,
                      "Moderate", "DiaBP", 79.19403, 0.8022114,
                      "Heavy", "SysBP", 135.7479, 0.9320387,
                      "Heavy", "DiaBP", 80.95062, 0.4976859)

# factoring data to ensure proper order
smokers_tb$smkerst <- factor(smokers_tb$smkerst, levels = c("Light", "Moderate", "Heavy"))


ggplot(smokers_tb,
       aes(x = BPtype, 
           y = value,
           fill = BPtype)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  geom_errorbar(aes(ymin = value - SE,
                    ymax = value + SE,
                    width = .75)) +
  facet_grid(~smkerst, switch = "x") +
  labs(title = "Effects of Smoking on Blood Pressure",
       subtitle = "Separated by Smoking Status",
       x = "Smoking Status",
       y = "Average Value (mmHG)",
       fill = "Type of Blood Pressure") +
  theme(axis.text.x = element_blank())
```

Bar graph for blood pressure seperated by sex:
```{r}
# creating new data frame to plot both SysBP and DiaBP into the same graph
smokers_tb_sex <- tribble(~smkerst, ~sex, ~BPtype, ~value, ~SE, 
                      "Light", "Male", "SysBP", 134.9444, 3.508537,
                      "Light", "Male", "DiaBP", 80.87500, 2.1920107,
                      "Light", "Female", "SysBP", 135.1316, 3.493090,
                      "Light", "Female", "DiaBP", 78.94737, 1.4470978,
                      "Moderate", "Male", "SysBP", 132.3014, 2.182780,
                      "Moderate", "Male", "DiaBP", 78.79452, 1.3806128,
                      "Moderate", "Female", "SysBP", 131.4102, 2.081844,
                      "Moderate", "Female", "DiaBP", 79.42188, 0.9869261,
                      "Heavy", "Male", "SysBP", 136.3764, 1.177746,
                      "Heavy", "Male", "DiaBP", 81.55535, 0.6434208, 
                      "Heavy", "Female", "SysBP",134.9558, 1.496394,
                      "Heavy", "Female", "DiaBP", 80.18837, 0.7782584)

# factoring data to ensure proper order
smokers_tb_sex$smkerst <- factor(smokers_tb_sex$smkerst, levels = c("Light", "Moderate", "Heavy"))


ggplot(smokers_tb_sex,
       aes(x = BPtype, 
           y = value,
           fill = BPtype)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  geom_errorbar(aes(ymin = value - SE,
                    ymax = value + SE,
                    width = .75)) +
  facet_grid(sex~smkerst, switch = "x") +
  labs(title = "Figure 1",
       subtitle = "The Effects of Smoking on Blood Pressure",
       x = "Smoking Status",
       y = "Average Value (mmHG)",
       fill = "Type of Blood Pressure") +
  theme(axis.text.x = element_blank())
```

Bar graph for heart rate:
```{r}
ggplot(smokers_descrip,
       aes(x = smkerst,
           y = Mean_Heartrate)) +
  geom_bar(stat = "identity",
           fill = "firebrick") +
  theme_minimal() +
  geom_errorbar(aes(ymin = Mean_Heartrate - SE_Heartrate,
                    ymax = Mean_Heartrate + SE_Heartrate,
                    width = .75)) +
  labs(title = "Effects of Smoking on Heart Rate",
       subtitle = "Separated by Smoking Status",
       x = "Smoking Status",
       y = "Average Value (bpm)")
```


Bar graph for Heart rate separated by sex:
```{r}
ggplot(smokers_descrip_sex,
       aes(x = smkerst,
           y = Mean_Heartrate,
           fill = sex)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  geom_errorbar(aes(ymin = Mean_Heartrate - SE_Heartrate,
                    ymax = Mean_Heartrate + SE_Heartrate,
                    width = .75)) +
  facet_wrap(~sex) +
  theme(legend.position = "none") +
  labs(title = "Figure 2",
       subtitle = "Effects of Smoking on Heart Rate",
       x = "Smoking Status",
       y = "Average Value (bpm)")
```

## ANOVA 

sysBP:
```{r}
smokers_sysbp_aov <- aov(sysbp ~ smkerst,
                   data = smokers)

summary(smokers_sysbp_aov)
```

diaBP:
```{r}
smokers_diabp_aov <- aov(diabp ~ smkerst,
                   data = smokers)

summary(smokers_diabp_aov)
```

heart rate:
```{r}
smokers_hr_aov <- aov(heartrte ~ smkerst,
                   data = smokers)

summary(smokers_hr_aov)
```



## Post-hocs

SysBP:
```{r}
ScheffeTest(smokers_sysbp_aov)
```


DiaBP:
```{r}
ScheffeTest(smokers_diabp_aov)
```




## ANOVA by sex

```{r}
fem = filter(smokers, sex == "Female")

head(fem)
```

Female sysbp:
```{r}
fem_sys_aov <- aov(sysbp ~ smkerst, 
               data = fem)

summary(fem_sys_aov)
```

Female diabp:
```{r}
fem_dia_aov <- aov(diabp ~ smkerst, 
               data = fem)

summary(fem_dia_aov)
```

```{r}
fem_hr_aov <- aov(heartrte ~ smkerst, 
               data = fem)

summary(fem_hr_aov)
```

Post Hoc:
```{r}
ScheffeTest(fem_sys_aov)
```




```{r}
mod = filter(smokers, smkerst == "Moderate")

head(mod)
```


Moderate by sex
```{r}
mod_hr_test <- t.test(heartrte ~ sex, 
                   data = mod, 
                   paired = F, 
                   alternative = "two.sided", 
                   var.equal = T, 
                   conf.level = .90)

mod_hr_test
```


